Carrier phase and amplitude estimation of $M$-ary phase shift keying using pilots and data

### 2012-05-17 paper.tex
added clt computing code.  A bit broken at the moment!
Added a stack of new simulations.  Fixed bug in clt code.  This now appears to work, but I would like to look at it a little closer.
Added to paper

With $M$-ary phase shift keying ($M$-PSK), the transmitted symbols take the form,
\[
s_i = e^{j u_i},
\]
where $j = \sqrt{-1}$ and $u_i$ is from the set $\{0, \tfrac{2\pi}{M}, \dots, \tfrac{2\pi(M-1)}{M}\}$ and $M$ is the size of the constellation.  We assume that some of the transmitted symbols are \emph{pilot symbols} known to the receiver and the remainder are information carrying \emph{data symbols} with phase that is unknown to the receiver.  So,
\[
s_i = \begin{cases}
p_i & i \in P, \\
d_i & i \in D, \\
0 & otherwise,
\end{cases}
\]
where $P$ is the set of indices describing the position of the pilot symbols $p_i$, and $D$ is a set of indices describing the position of the data symbols $d_i$.  The sets $P$ and $D$ are disjoint, i.e. $P \cap D = \emptyset$.

We assume that time offset estimation (symbol timing) and matched filtering have already been performed.  The recieved symbols are then given by,
\begin{equation}\label{eq:sigmod}
y_i = a_0 s_i + w_i, \qquad i \in P \cup D,
\end{equation}
where $w_i$ is noise and $a_0 = \rho_0 e^{j\theta_0}$ is a complex number representing both carrier phase $\theta_0$ and amplitude $\rho_0$ (by definition $\rho_0$ is a positive real number).  Our aim is to estimate $a_0$ from the noisy symbols $\{ y_i, i \in P \cup D \}$.  Complicating matters is that the data symbols $d_i$ are not known to the reciever and must also be estimated.  Estimation problems of this type have undergone extensive prior study~\cite{Wilson1989,Makrakis1990,Liu1991,Mackenthun1994,Sweldens2001}.  A practical approach is the least squares estimator, that is, the minimisers of the sum of squares function
\begin{align*}
SS(a, &\{d_i, i \in D\}) = \sum_{i \in P \cup D} \abs{ y_i - a s_i }^2 \\
&= \sum_{i \in P} \abs{ y_i - a s_i }^2 + \sum_{i \in P \cup D} \abs{ y_i - a d_i }^2,
\end{align*}
where $\abs{x}$ denotes the magnitude of the complex number $x$.  The least squares estimator is also the maximum likelihood estimator under the assumption that the noise sequence $\{w_i, i \in \ints\}$ is additive white and Gaussian.  However, as we will show, the estimator works well under less stringent assumptions.  %It is the least squares estimator that we primarily study in this paper.

The existing literature mostly considers what is called \emph{noncoherent detection} where no pilots symbols exist ($P = \emptyset$ where $\emptyset$ is the empty set).  In the nocoherent setting \emph{differential encoding} is often used, and for this reason the estimation problem has been called \emph{multiple symbol differential detection}.  An important paper is by Mackenthun~\cite{Mackenthun1994} who described an algorithm to compute the least squares estimator requiring only $O(L \log L)$ arithmetic operations, where $L = \abs{P \cup D}$ is the total number of symbols transmitted.  Sweldens~\cite{Sweldens2001} rediscovered Mackenthun's algorithm in 2001.  Here, we consider a slightly more general model than is usual, one that allows the utilisation of pilot symbols, i.e., \emph{coherent detection}.  Our model includes the noncoherent case by setting the number of pilot symbols to zero, that is, putting $P = \emptyset$.  We show in Section~\ref{sec:least-squar-estim}~that only a minor modification of Mackenthun's algorithm is required to include pilot symbols.

@@

Define the sum of squares function
\begin{align}
SS(a, &\{d_i, i \in D\}) = \sum_{i \in P \cup D} \abs{ y_i - a s_i }^2 \nonumber \\
&= \sum_{i \in P \cup D} \abs{y_i}^2 - a s_i y_i^* - a^* s_i^* y_i + aa^*. \label{eq:SS}
\end{align}
Fixing the data symbols $\{d_i, i \in D\}$, differentiating with respect to $a^*$, and setting the resulting equation to zero we find that the least squares estimator of $a_0$, satisfies
\begin{equation}\label{eq:hata}
\hat{a} = \frac{1}{L} \sum_{i \in P \cup D} y_i s_i^* = \frac{1}{L} Y
\end{equation}
where $L = \abs{P \cup D}$ is the total number of symbols transmitted, and to simplify our notation we have put 
\[
Y = \sum_{i \in P \cup D} y_i s_i^* = \sum_{i \in P } y_i p_i^* + \sum_{i \in D } y_i d_i^*.
\]  
Note that $Y$ is a function of the unknown data symbols $\{ d_i, i \in D\}$ and we could write $Y(\{ d_i, i \in D\})$, but have chosen to surpress the argument $(\{ d_i, i \in D\})$ for notational clarity.  Substituting $\hat{a} = \frac{1}{L}Y$ into~\eqref{eq:SS} we obtain the sum of squares function, conditioned on minimisation with respect to $a$,
\begin{equation}\label{eq:SSdatasymbols}
SS(\{d_i, i \in D\}) = A - \frac{1}{L}\abs{Y}^2,
\end{equation}
where $A = \sum_{i \in P \cup D}\abs{y_i}^2$ is a constant.  Observe that given a candidate value for the data symbols, we can compute the corresponding sum of squares function $SS(\{d_i, i \in D\})$ in $O(L)$ arithmetic operations.  It turns out that there are atmost $(M+1)\abs{D}$ candidate values of the least squares estimator of the data symbols~\cite{Sweldens2001,Mackenthun1994}.  %When the number of data symbols is not small, this set is substantially smaller than the entire set of possible transmitted symbols, which is of size $M^{\abs{D}}$.

To see this, let $a = \rho e^{j\theta}$ where $\rho$ is a nonnegative real.  Now the sum of squares function can be written as
\begin{align}
SS(\rho, \theta, &\{d_i, i \in D\}) = \sum_{i \in P \cup D} \abs{ y_i - \rho e^{j\theta} s_i }^2  \nonumber \\
&= \sum_{i \in P} \abs{ y_i - \rho e^{j\theta} p_i }^2 + \sum_{i \in D} \abs{ y_i - \rho e^{j\theta} d_i }^2. \label{eq:SSallparams}
\end{align}
For given $\theta$, the least squares estimator of the $i$th data symbol $d_i$ is given by minimising $\abs{ y_i - \rho e^{j\theta} d_i }^2$, that is,
\begin{equation}\label{eq:hatdfinxtheta}
\hat{d}_i(\theta) = e^{j\hat{u}_i(\theta)} \qquad \text{where} \qquad \hat{u}_i(\theta) = \round{\angle( e^{-j\theta}y_i)},
\end{equation}
and where $\angle(\cdot)$ denotes the complex argument (or phase), and $\round{\cdot}$ rounds its argument to the nearest multiple of $\frac{2\pi}{M}$.  A word of caution, it is often the case that $\round{\cdot}$ denotes rounding to the nearest \emph{integer}.  This is not the case here.  If the function $\operatorname{round}(\cdot)$ takes its argument to the nearest integer then,
\[
\round{x} = \tfrac{2\pi}{M}\operatorname{round}\left(\tfrac{M}{2\pi}x\right).
\] 
Note that $\hat{d}_i(\theta)$ does not depend on the amplitude $\rho$.  As defined, $\hat{u}_i(\theta)$ is not strictly inside the set $\{0, \tfrac{2\pi}{M}, \dots, \tfrac{2\pi(M-1)}{M}\}$, but this is not of consequence, as we intend its value to be considered equivalent modulo $2\pi$.  With this in mind,
\[
\hat{u}_i(\theta) = \round{\angle{y_i} - \theta }
\]
which is equivalent to the definition from~\eqref{eq:hatdfinxtheta} modulo $2\pi$.

@@

Let $S$ be the set of all $f(\theta)$ as $\theta$ varies from $0$ to $2\pi$, that is,
\[
S = \{ f(\theta) \mid \theta \in [0, 2 \pi) \}.
\]
If $\hat{\theta}$ is the least squares estimator the phase, then it is clear from~\eqref{eq:hatdfinxtheta}, that $S$ contains the sequence $\{ \hat{d}_i(\hat{\theta}), i \in D \}$ corresponding to the least squares estimator of the data symbols, i.e., $S$ contains the minimiser of~\eqref{eq:SSdatasymbols}.  Observe from~\eqref{eq:uicombos} that there are atmost $(M+1)\abs{D}$ sequences in $S$, because there are $M+1$ sequences for each $i \in D$.

@@

Let,
\begin{equation}\label{eq:SSfk}
SS(f_k) = A - \frac{1}{L}\abs{Y_k}^2,
\end{equation}
where, 
\begin{align*}
Y_k = Y( f_k ) &= \sum_{i \in P} y_i p_i^*  + \sum_{i \in D} y_i f_{ki}^* \\
&= B + \sum_{i \in D}g_{ki},
\end{align*}
where $B = \sum_{i \in P} y_i p_i^*$ is a constant, independent of the data symbols, and $f_{ki}$ denotes the $i$th symbol in the sequence $f_k$, and for convenience, we put $g_{ki}  = y_i f_{ki}^*$.  Letting $g_{k}$ be the sequence $\{ y_i f_{ki}^*, i \in D\}$ we have, from~\eqref{eq:fkrec}, that $g_k$ satisfies the recursive equation
\[
g_{k+1} = g_{k} e_{\sigma(k)}^*,
\]
where $g_{k} e_{\sigma(k)}^*$ indicates the sequence $g_k$ with the $\sigma(k)$th element replaced by $g_{k \sigma(k)}e^{j2\pi/M}$.  Now,
\[
Y_0 = B + \sum_{i \in D} g_{0i}
\] 
can be computed in $O(L)$ operations, and
\begin{align*}
Y_1 &= B + \sum_{i \in D} g_{1i} \\
&= B +  (e^{j2\pi/M} - 1)g_{0\sigma(0)} + \sum_{i \in D} g_{0i} \\
&= Y_0 + \eta g_{0\sigma(0)},
\end{align*}
where $\eta = (e^{j2\pi/M} - 1)$.  In general we have,
\[
Y_{k+1} = Y_k + \eta g_{k\sigma(k)}.
\]
So, each $Y_k$ can be computed from it predecessor $Y_{k-1}$ in a constant number of operations.  Given $Y_k$, the value of $SS(f_k)$ can be computed in a constant number of operations using~\eqref{eq:SSfk}.  Let $\hat{k}$ be the integer that minimises $SS(f_k)$, i.e. $\hat{k} = \arg\min SS(f_k)$.  The least squares estimator of the complex amplitude is then computed according to~\eqref{eq:hata}, as 
\begin{equation}\label{eq:ahatYhat}
\hat{a} = \frac{1}{L} Y_{\hat{k}}.
\end{equation}
Psuedocode is given in Algorithm~\ref{alg:loglinear}.  Line~\ref{alg_sortindices} contains the function $\operatorname{sortindicies}$ that, given the sequence $z = \{z_i, i \in D\}$, returns the permutation $\sigma$ of the indices in $D$ as described in~\eqref{eq:sigmasortind}.  Computing the $\operatorname{sortindicies}$ function requires sorting $\abs{D}$ elements.  This requires $O(L \log L)$ operations.  The $\operatorname{sortindicies}$ function is the primary bottleneck in this algorithm when $L$ is large.  The loops on lines~\ref{alg_loop_setup} and~\ref{alg_loop_search} and the operations on lines~\ref{alg_Y} to lines~\ref{alg_Q} all require $O(L)$ or less operations.  %In the next sections we will show how to the sortinces function can be avoided.  This leads to an algorithm that requires only $O(L)$ operations.

@@

\section{Circularly symmetric complex random variables}

Before describing the statistical properties of the least squares estimator, we first require some properties of complex valued random variables.  A complex random variable $X$ is called \emph{circularly symmetric} if the distribution of its phase $\angle{X}$ is uniform on $[0,2\pi)$ and is independent of the distribution of its magnitude $\abs{X}$.  That is, if we define real random variables $Z > 0$ and $\Theta \in [0,2\pi)$ such that $Ze^{j\Theta} = X$, then $\Theta$ is uniformly distributed on $[0,2\pi)$ and is indepedent of $Z$.  If the probability density function of $Z$ is $f_Z(z)$, then the joint density function of $\Theta$ and $Z$ is 
\[
f_{Z,\Theta}(z,\theta) = \frac{1}{2\pi}f_Z(z).
\]  
Observe that for any real number $\phi$, the distribution of $X$ and $e^{j\phi}X$ are the same, that is, the distribution is invariate to phase rotation.  If $\expect\abs{X} = \expect Z$ is finite, then $X$ has zero mean since
\begin{align*}
\expect X &= \int_{0}^{2\pi} \int_{0}^\infty z e^{j\theta} f_{Z,\Theta}(z,\theta) dz d\theta \\
&= \frac{1}{2\pi} \int_{0}^{2\pi} e^{j\theta} \int_{0}^\infty z f_Z(z) dz d\theta \\
&= \frac{1}{2\pi}\expect Z \int_{0}^{2\pi} e^{j\theta} d\theta = 0.
\end{align*}

We will have particular use of complex random variables of the form $1 + X$.  Let $R \geq 0$ and $\Phi \in [0,2\pi)$ be real random variables satisfying, 
\[
R e^{j\Phi} = 1 + X.
\]
The joint distribution of $R$ and $\Phi$ can be show to be
\[
f(r,\phi) = \frac{r f_Z(\sqrt{r^2 - 2r\cos\phi + 1})}{\sqrt{r^2 - 2r\cos\phi + 1}}.
\]
The mean of $R e^{j\Phi}$ is equal to one because the mean of $X$ is zero.  So,
\[
\expect \Re(R e^{j\Phi}) = \expect R \cos(\Phi) = 1,
\]
where $\Re(\cdot)$ denotes the real part, and
\[
\expect \Im(R e^{j\Phi}) = \expect R \sin(\Phi) = 0,
\]
where $\Im(\cdot)$ denotes the imaginary part.

@@

\begin{theorem}\label{thm:normality} (Asymptotic normality)
Under the same conditions as Theorem~\ref{thm:consistency}, let $f(r,\phi)$ be the joint probability density function of $R_1$ and $\Phi_1$, and let
\[
g(\phi) = \int_{0}^{\infty} r f(r,\phi) dr.
\]
Put $\lambda_L = \sfracpart{\theta_0 - \hat{\theta}}_\pi$ and $m_L = \hat{\rho} - \rho_0 G(0)$. If $g(\phi)$ is continuous at $\phi = (2k+1)\tfrac{\pi}{M}$ for each $k = 0, 1, \dots M-1$, then the distribution of $(\sqrt{L}\lambda_L, \sqrt{L}m_L)$ converges to the bivariate normal with zero mean and covariance
\[
C = \frac{1}{(p + H d)^2} \left( \begin{array}{cc} 
A_1 + A_2 & ? \\
? & ?
\end{array} \right)
\]
as $L \rightarrow \infty$, where
\[
H = 2 \sin(\tfrac{\pi}{M}) \sum_{k = 0}^{M-1} g(\tfrac{(2k+1)\pi}{M}),
\]
\[
A_1 = \expect R_1^2\sin^2(\Phi_1), \qquad A_2 = \expect R_1^2\sin^2\fracpart{\Phi_1}.
\]
\end{theorem}

@@

Proofs of these theorems are given in the follows subsections~\ref{sec:proof-almost-sure}~and~\ref{sec:proof-asympt-norm}.  We first discussion the assumptions made by these theorems.

\begin{lemma}
Let $X = Z e^{j\Theta}$ be a circularly symmetric complex random variable with pdf $\tfrac{1}{2\pi}f_Z(z)$ such that $z^{-1} f_Z(z)$ is nonincreasing with $z$.  Let $R > 0$ and $\Phi \in [0, 2\pi)$ be random variables satisfying $R e^{j\Phi} = 1 + X$.  Let $h_2$ and $k_2$ be functions 
\[
h_2(x) = \expect R \cos\sfracpart{x + \Phi}, \;\;\; k_2(x) =  \expect R \sin\sfracpart{ x + \Phi}.
\]
Then $h_2(x)$ is uniquely maximised at $x=0$ other the interval $[-\tfrac{\pi}{M},\tfrac{\pi}{M})$ and $k_2(0) = 0$.
\end{lemma}

@@

\begin{lemma}\label{lem:convtoexpGlamL}
\[
\prob\{ \sabs{G(\hat{\lambda}_L) - G(0)} > \epsilon \} = O(e^{-\epsilon^2 L})
\]
\end{lemma}
\begin{IEEEproof}
Lemma~\ref{lem:Zconunif} shows BLERG that 
\[
0 \leq G(\hat{\lambda}_L) - G(0),
\]
and since $\hat{\lambda}_L$ is the maximiser of $G_L$ then
\[ 
0 \leq G_L(0) - G_L(\hat{\lambda}_L).
\]
Adding both inequalities and taking appropriate absolute values gives
\begin{align*}
0 &\leq G(\hat{\lambda}_L) - G(0) \\ 
& \leq G(\hat{\lambda}_L) - G(0) + G_L(0) - G_L(\hat{\lambda}_L) \\
& \leq |G(\hat{\lambda}_L) - G_L(\hat{\lambda}_L)| + |G_L(0) - G(0)|,
\end{align*}
BLERG
\end{IEEEproof}

>>>>

With $M$-ary phase shift keying ($M$-PSK), the transmitted symbols take the form,
\[
s_i = e^{j u_i},
\]
where $j = \sqrt{-1}$ and $u_i$ is from the set $\{0, \tfrac{2\pi}{M}, \dots, \tfrac{2\pi(M-1)}{M}\}$ and $M$ is the size of the constellation.  We assume that some of the transmitted symbols are \emph{pilot symbols} known to the receiver and the remainder are information carrying \emph{data symbols} with phase that is unknown to the receiver.  So,
\[
s_i = \begin{cases}
p_i & i \in P \\
d_i & i \in D,
\end{cases}
\]
where $P$ is the set of indices describing the position of the pilot symbols $p_i$, and $D$ is a set of indices describing the position of the data symbols $d_i$.  The sets $P$ and $D$ are disjoint, i.e. $P \cap D = \emptyset$, and $L = \abs{P \cup D}$ is the total number of symbols transmitted.

We assume that time offset estimation (symbol timing) and matched filtering have already been performed.  The recieved symbols are then given by,
\begin{equation}\label{eq:sigmod}
y_i = a_0 s_i + w_i, \qquad i \in P \cup D,
\end{equation}
where $w_i$ is noise and $a_0 = \rho_0 e^{j\theta_0}$ is a complex number representing both carrier phase $\theta_0$ and amplitude $\rho_0$ (by definition $\rho_0$ is a positive real number).  Our aim is to estimate $a_0$ from the noisy symbols $\{ y_i, i \in P \cup D \}$.  Complicating matters is that the data symbols $d_i$ are not known to the reciever and must also be estimated.  Estimation problems of this type have undergone extensive prior study~\cite{ViterbiViterbi_phase_est_1983,Cowley_ref_sym_carr_1998,Wilson1989,Makrakis1990,Liu1991,Mackenthun1994,Sweldens2001}.  A practical approach is the least squares estimator, that is, the minimisers of the sum of squares function
\begin{align*}
SS(a, &\{d_i, i \in D\}) = \sum_{i \in P \cup D} \abs{ y_i - a s_i }^2 \\
&= \sum_{i \in P} \abs{ y_i - a s_i }^2 + \sum_{i \in P \cup D} \abs{ y_i - a d_i }^2,
\end{align*}
where $\abs{x}$ denotes the magnitude of the complex number $x$.  The least squares estimator is also the maximum likelihood estimator under the assumption that the noise sequence $\{w_i, i \in \ints\}$ is additive white and Gaussian.  However, as we will show, the estimator works well under less stringent assumptions.  %It is the least squares estimator that we primarily study in this paper.

The existing literature mostly considers what is called \emph{noncoherent detection} where no pilots symbols exist ($P = \emptyset$ where $\emptyset$ is the empty set).  In the nocoherent setting \emph{differential encoding} is often used, and for this reason the estimation problem has been called \emph{multiple symbol differential detection}.  A popular approach is the so called \emph{non-data aided}, sometimes also called \emph{non-decision directed}, estimator based on the paper of Viterbi and Viterbi~\cite{ViterbiViterbi_phase_est_1983}.  The idea is the `strip' the modulation from the recieved signal by taking $y_i / \abs{y_i}$ to the power of $M$.  A function $F$, mapping $\reals$ to $\reals$, is chosen and the estimator of the carrier phase $\theta_0$ is taken to be $\tfrac{1}{M}\angle{A}$ where $\angle$ denotes the complex argument and $A$ is the average
\[
A = \frac{1}{L}\sum_{i \in P \cup D} F(\abs{y_i}) \big(\tfrac{y_i}{\abs{y_i}}\big)^M.
\]
Various choices for $F$ are suggested in~\cite{ViterbiViterbi_phase_est_1983} and a statistical analysis is presented.  A caveat of this estimator is that it is not entirely obvious how pilot symbols should be included, so as to made the estimator coherent. Techniques are available but they are somewhat adhoc~\cite{Cowley_ref_sym_carr_1998}.  This problem does not occur with the least square estimator.

An important paper is by Mackenthun~\cite{Mackenthun1994} who described an algorithm to compute the least squares estimator requiring only $O(L \log L)$ arithmetic operations.  Sweldens~\cite{Sweldens2001} rediscovered Mackenthun's algorithm in 2001.  Both Mackenthun and Swelden's considered the noncoherent setting, but we show in Section~\ref{sec:least-squar-estim}~that only a minor modification of Mackenthun's algorithm is required to include pilot symbols. Our model includes the noncoherent case by setting the number of pilot symbols to zero, that is, putting $P = \emptyset$.

@@

Define the sum of squares function
\begin{align}
SS(a, &\{d_i, i \in D\}) = \sum_{i \in P \cup D} \abs{ y_i - a s_i }^2 \nonumber \\
&= \sum_{i \in P \cup D} \abs{y_i}^2 - a s_i y_i^* - a^* s_i^* y_i + aa^*, \label{eq:SS}
\end{align}
where $*$ denote the complex congugate.  Fixing the data symbols $\{d_i, i \in D\}$, differentiating with respect to $a^*$, and setting the resulting equation to zero we find that the least squares estimator of $a_0$, satisfies
\begin{equation}\label{eq:hata}
\hat{a} = \frac{1}{L} \sum_{i \in P \cup D} y_i s_i^* = \frac{1}{L} Y
\end{equation}
where $L = \abs{P \cup D}$ is the total number of symbols transmitted, and to simplify our notation we have put 
\[
Y = \sum_{i \in P \cup D} y_i s_i^* = \sum_{i \in P } y_i p_i^* + \sum_{i \in D } y_i d_i^*.
\]  
Note that $Y$ is a function of the unknown data symbols $\{ d_i, i \in D\}$ and we could write $Y(\{ d_i, i \in D\})$, but have chosen to surpress the argument $(\{ d_i, i \in D\})$ for notational clarity.  Substituting $\hat{a} = \frac{1}{L}Y$ into~\eqref{eq:SS} we obtain the sum of squares function, conditioned on minimisation with respect to $a$,
\begin{equation}\label{eq:SSdatasymbols}
SS(\{d_i, i \in D\}) = A - \frac{1}{L}\abs{Y}^2,
\end{equation}
where $A = \sum_{i \in P \cup D}\abs{y_i}^2$ is a constant.  Observe that given a candidate value for the data symbols, we can compute the corresponding sum of squares function $SS(\{d_i, i \in D\})$ in $O(L)$ arithmetic operations.  It turns out that there are atmost $(M+1)\abs{D}$ candidate values of the least squares estimator of the data symbols~\cite{Sweldens2001,Mackenthun1994}.  %When the number of data symbols is not small, this set is substantially smaller than the entire set of possible transmitted symbols, which is of size $M^{\abs{D}}$.

To see this, let $a = \rho e^{j\theta}$ where $\rho$ is a nonnegative real.  Now the sum of squares function can be written as
\begin{align}
SS(\rho, \theta, &\{d_i, i \in D\}) = \sum_{i \in P \cup D} \abs{ y_i - \rho e^{j\theta} s_i }^2  \nonumber \\
&= \sum_{i \in P} \abs{ y_i - \rho e^{j\theta} p_i }^2 + \sum_{i \in D} \abs{ y_i - \rho e^{j\theta} d_i }^2. \label{eq:SSallparams}
\end{align}
For given $\theta$, the least squares estimator of the $i$th data symbol $d_i$ is given by minimising $\abs{ y_i - \rho e^{j\theta} d_i }^2$, that is,
\begin{equation}\label{eq:hatdfinxtheta}
\hat{d}_i(\theta) = e^{j\hat{u}_i(\theta)} \qquad \text{where} \qquad \hat{u}_i(\theta) = \round{\angle( e^{-j\theta}y_i)},
\end{equation}
and where $\angle(\cdot)$ denotes the complex argument (or phase), and $\round{\cdot}$ rounds its argument to the nearest multiple of $\frac{2\pi}{M}$.  A word of caution, the notation $\round{\cdot}$ is often used to denote rounding to the nearest \emph{integer}.  This is not the case here.  If the function $\operatorname{round}(\cdot)$ takes its argument to the nearest integer then,
\[
\round{x} = \tfrac{2\pi}{M}\operatorname{round}\left(\tfrac{M}{2\pi}x\right).
\] 
Note that $\hat{d}_i(\theta)$ does not depend on $\rho$.  As defined, $\hat{u}_i(\theta)$ is not strictly inside the set $\{0, \tfrac{2\pi}{M}, \dots, \tfrac{2\pi(M-1)}{M}\}$, but this is not of consequence, as we intend its value to be considered equivalent modulo $2\pi$.  With this in mind,
\[
\hat{u}_i(\theta) = \round{\angle{y_i} - \theta }
\]
which is equivalent to the definition from~\eqref{eq:hatdfinxtheta} modulo $2\pi$.

@@

Let $S$ be the set of all sequences $f(\theta)$ as $\theta$ varies from $0$ to $2\pi$, that is,
\[
S = \{ f(\theta) \mid \theta \in [0, 2 \pi) \}.
\]
If $\hat{\theta}$ is the least squares estimator the phase, then it follows from~\eqref{eq:hatdfinxtheta}, that $S$ contains the sequence $\{ \hat{d}_i(\hat{\theta}), i \in D \}$ corresponding to the least squares estimator of the data symbols, i.e., $S$ contains the minimiser of~\eqref{eq:SSdatasymbols}.  Observe from~\eqref{eq:uicombos} that there are atmost $(M+1)\abs{D}$ sequences in $S$, because there are $M+1$ sequences for each $i \in D$.

@@

Let,
\begin{equation}\label{eq:SSfk}
SS(f_k) = A - \frac{1}{L}\abs{Y_k}^2,
\end{equation}
where, 
\begin{align*}
Y_k = Y( f_k ) &= \sum_{i \in P} y_i p_i^*  + \sum_{i \in D} y_i f_{ki}^* \\
&= B + \sum_{i \in D}g_{ki},
\end{align*}
where $B = \sum_{i \in P} y_i p_i^*$ is a constant, independent of the data symbols, and $f_{ki}$ denotes the $i$th symbol in the sequence $f_k$, and for convenience, we put $g_{ki}  = y_i f_{ki}^*$.  Letting $g_{k}$ be the sequence $\{g_{ik}, i \in D\}$ we have, from~\eqref{eq:fkrec}, that $g_k$ satisfies the recursive equation
\[
g_{k+1} = g_{k} e_{\sigma(k)}^*,
\]
where $g_{k} e_{\sigma(k)}^*$ indicates the sequence $g_k$ with the $\sigma(k)$th element replaced by $g_{k \sigma(k)}e^{j2\pi/M}$.  Now,
\[
Y_0 = B + \sum_{i \in D} g_{0i}
\] 
can be computed in $O(L)$ operations, and
\begin{align*}
Y_1 &= B + \sum_{i \in D} g_{1i} \\
&= B +  (e^{j2\pi/M} - 1)g_{0\sigma(0)} + \sum_{i \in D} g_{0i} \\
&= Y_0 + \eta g_{0\sigma(0)},
\end{align*}
where $\eta = (e^{j2\pi/M} - 1)$.  In general we have,
\[
Y_{k+1} = Y_k + \eta g_{k\sigma(k)}.
\]
So, each $Y_k$ can be computed from it predecessor $Y_{k-1}$ in a constant number of operations.  Given $Y_k$, the value of $SS(f_k)$ can be computed in a constant number of operations using~\eqref{eq:SSfk}.  Let $\hat{k}$ be the integer that minimises $SS(f_k)$, i.e. $\hat{k} = \arg\min SS(f_k)$.  The least squares estimator of the complex amplitude is then computed according to~\eqref{eq:hata}, as 
\begin{equation}\label{eq:ahatYhat}
\hat{a} = \frac{1}{L} Y_{\hat{k}}.
\end{equation}
Psuedocode is given in Algorithm~\ref{alg:loglinear}.  Line~\ref{alg_sortindices} contains the function $\operatorname{sortindicies}$ that, given the sequence $z = \{z_i, i \in D\}$, returns the permutation $\sigma$ of the indices in $D$ as described in~\eqref{eq:sigmasortind}.  The $\operatorname{sortindicies}$ function requires sorting $\abs{D}$ elements.  This requires $O(L \log L)$ operations.  The $\operatorname{sortindicies}$ function is the primary bottleneck in this algorithm when $L$ is large.  The loops on lines~\ref{alg_loop_setup} and~\ref{alg_loop_search} and the operations on lines~\ref{alg_Y} to lines~\ref{alg_Q} all require $O(L)$ or less operations.  %In the next sections we will show how to the sortinces function can be avoided.  This leads to an algorithm that requires only $O(L)$ operations.

@@

\section{Circularly symmetric complex random variables}\label{sec:circ-symm-compl}

Before describing the statistical properties of the least squares estimator, we first require some properties of complex valued random variables.  A complex random variable $X$ is said to be \emph{circularly symmetric} if the distribution of its phase $\angle{X}$ is uniform on $[0,2\pi)$ and is independent of the distribution of its magnitude $\abs{X}$.  That is, if we define real random variables $Z > 0$ and $\Theta \in [0,2\pi)$ such that $Ze^{j\Theta} = X$, then $\Theta$ is uniformly distributed on $[0,2\pi)$ and is indepedent of $Z$.  If the probability density function of $Z$ is $f_Z(z)$, then the joint density function of $\Theta$ and $Z$ is 
\[
f_{Z,\Theta}(z,\theta) = \frac{1}{2\pi}f_Z(z).
\]  
Observe that for any real number $\phi$, the distribution of $X$ and $e^{j\phi}X$ are the same, that is, the distribution is invariant to phase rotation.  If $\expect\abs{X} = \expect Z$ is finite, then $X$ has zero mean since
\begin{align*}
\expect X &= \int_{0}^{2\pi} \int_{0}^\infty z e^{j\theta} f_{Z,\Theta}(z,\theta) dz d\theta \\
&= \frac{1}{2\pi} \int_{0}^{2\pi} e^{j\theta} \int_{0}^\infty z f_Z(z) dz d\theta \\
&= \frac{1}{2\pi}\expect Z \int_{0}^{2\pi} e^{j\theta} d\theta = 0.
\end{align*}

We will have particular use of complex random variables of the form $1 + X$.  Let $R \geq 0$ and $\Phi \in [0,2\pi)$ be real random variables satisfying, 
\[
R e^{j\Phi} = 1 + X.
\]
The joint distribution of $R$ and $\Phi$ can be shown to be
\[
f(r,\phi) = \frac{r f_Z(\sqrt{r^2 - 2r\cos\phi + 1})}{\sqrt{r^2 - 2r\cos\phi + 1}}.
\]
The mean of $R e^{j\Phi}$ is equal to one because the mean of $X$ is zero.  So,
\[
\expect \Re(R e^{j\Phi}) = \expect R \cos(\Phi) = 1,
\]
where $\Re(\cdot)$ denotes the real part, and
\[
\expect \Im(R e^{j\Phi}) = \expect R \sin(\Phi) = 0,
\]
where $\Im(\cdot)$ denotes the imaginary part.  The next lemma will be useful for the analysis of the least squares estimator.

\begin{lemma}\label{lem:h1minedcircsym}
Let $X = Z e^{j\Theta}$ be a circularly symmetric complex random variable with pdf $\tfrac{1}{2\pi}f_Z(z)$ such that $z^{-1} f_Z(z)$ is nonincreasing with $z$.  Let $R > 0$ and $\Phi \in [0, 2\pi)$ be random variables satisfying $R e^{j\Phi} = 1 + X$.  Let $h_1(x)$ be the function
\[
h_1(x) = \expect R \cos(x + \Phi).
\]
Then $h_1(x)$ is uniquley maximised as $0$ over the interval $[-\pi,\pi)$.
\end{lemma}
\begin{IEEEproof}
BLERG
\end{IEEEproof}

@@

\begin{theorem}\label{thm:normality} (Asymptotic normality)
Under the same conditions as Theorem~\ref{thm:consistency}, let $f(r,\phi)$ be the joint probability density function of $R_1$ and $\Phi_1$, and let
\[
g(\phi) = \int_{0}^{\infty} r f(r,\phi) dr.
\]
Put $\hat{\lambda}_L = \sfracpart{\theta_0 - \hat{\theta}}_\pi$ and $\hat{m}_L = \hat{\rho} - \rho_0 G(0)$. If $g(\phi)$ is continuous at $\phi = (2k+1)\tfrac{\pi}{M}$ for each $k = 0, 1, \dots M-1$, then the distribution of $(\sqrt{L}\lambda_L, \sqrt{L}m_L)$ converges to the bivariate normal with zero mean and covariance
\[
C = \frac{1}{(p + H d)^2} \left( \begin{array}{cc} 
pA_1 + dA_2 & ? \\
? & ?
\end{array} \right)
\]
as $L \rightarrow \infty$, where
\[
H = h_2(0) -  2 \sin(\tfrac{\pi}{M}) \sum_{k = 0}^{M-1} g(\tfrac{(2k+1)\pi}{M}),
\]
\[
A_1 = \expect R_1^2\sin^2(\Phi_1), \qquad A_2 = \expect R_1^2\sin^2\fracpart{\Phi_1}.
\]
\end{theorem}

@@

Proofs of these theorems are given in subsections~\ref{sec:proof-almost-sure}~and~\ref{sec:proof-asympt-norm}.  We first discuss the assumptions made by these theorems.

\begin{lemma}
Under the same conditions as Lemma~\ref{lem:h1minedcircsym}, define the functions $h_2$ and $k_2$, 
\[
h_2(x) = \expect R \cos\sfracpart{x + \Phi}, \;\;\; k_2(x) =  \expect R \sin\sfracpart{ x + \Phi}.
\]
Then $h_2(x)$ is uniquely maximised at $x=0$ other the interval $[-\tfrac{\pi}{M},\tfrac{\pi}{M})$ and $k_2(0) = 0$.
\end{lemma}

@@

\begin{lemma}\label{lem:convtoexpGlamL} For any $\epsilon > 0$,
\[
\prob\{ \sabs{G(\hat{\lambda}_L) - G(0)} > \epsilon \} = O(e^{-\epsilon^2 L})
\]
\end{lemma}
\begin{IEEEproof}
BLERG
% Lemma~\ref{lem:Zconunif} shows BLERG that 
% \[
% 0 \leq G(\hat{\lambda}_L) - G(0),
% \]
% and since $\hat{\lambda}_L$ is the maximiser of $G_L$ then
% \[ 
% 0 \leq G_L(0) - G_L(\hat{\lambda}_L).
% \]
% Adding both inequalities and taking appropriate absolute values gives
% \begin{align*}
% 0 &\leq G(\hat{\lambda}_L) - G(0) \\ 
% & \leq G(\hat{\lambda}_L) - G(0) + G_L(0) - G_L(\hat{\lambda}_L) \\
% & \leq |G(\hat{\lambda}_L) - G_L(\hat{\lambda}_L)| + |G_L(0) - G(0)|,
% \end{align*}
\end{IEEEproof}

% \begin{lemma}\label{lem:Zconunif}
% \[
% \prob\left\{\sup_{\lambda \in [-\pi, \pi)}\abs{G_L(\lambda) - G(\lambda)} > \epsilon \right\} = O(e^{-\epsilon^2 L}).
% \]
% \end{lemma}
% \begin{IEEEproof}
% \end{IEEEproof}
